An Efficient Interpolation Technique for Jump Proposals in Reversible-Jump Markov Chain Monte Carlo Calculations

### 2010-12-19 paper.tex
Added references in section 2, first sentence in abstract, responded to Will's section 2 comments.

@@

\begin{abstract}
  \ilya{Add physics/astro motivation} Reversible-jump Markov chain
  monte carlo (RJMCMC) is an extremely powerful technique for
  performing Bayesian model selection, but it suffers from a
  fundamental difficulty: it requires jumps between model parameter
  spaces, but cannot retain a memory of the favored locations in more
  than one parameter space at a time.  Thus, a naive jump between
  parameter spaces is unlikely to be accepted in the MCMC algorithm
  and convergence is correspondingly slow.  Here we demonstrate an
  interpolation technique that uses samples from single-model MCMCs to
  propose inter-model jumps from an approximation to the single-model
  posterior of the target parameter space.  The interpolation
  technique, based on a kD-tree data structure, is adaptive and
  efficient in arbitrary dimensions.  We show that our technique leads
  to dramatically improved convergence over naive jumps in an RJMCMC,
  and compare it to other proposals in the literature to improve the
  convergence of RJMCMCs.  We also discuss the use of the same
  interpolation technique in two other contexts: as a convergence test
  for a single-model MCMC and as a way to construct efficient
  ``global'' proposal distributions for single-model MCMCs without
  prior knowledge of the structure of the posterior distribution.
\end{abstract}

@@

When selecting among alternative models, we are interested in finding
the model with the highest posterior probability $p(M_i|d)$.  However,
attempts to directly compute the evidence by performing the
integration in Eq.~\eqref{evidence} are generally very difficult in a
multi-dimensional, multi-modal parameter space when the likelihood has
to be evaluated numerically.  In particular, a grid-based integral
quickly becomes computationally unfeasible as the dimensionality of
$\vtheta$ exceeds a few.  The parameter space must typically be
explored in a stochastic manner before the evidence integral can be
computed.  There are several stochastic parameter-exploration
techniques focused directly on evidence computation (e.g., nested
sampling and MultiNest \ilya {Add references}).  However, we
frequently want to compute the posterior PDFs within each model along
with the evidences for the various models.  One of the most common
techniques for computing posterior PDFs in the context of a model is
the Markov Chain Monte Carlo, which we now describe.

@@

\will{Here $i$ should become $j$ (or some other index), because we
  have used $i$ already to denote the model.}

A Markov chain Monte Carlo produces a set of samples $\{ \vtheta^{(i)} \, | \, i = 1, \ldots \}$ from the model parameter space that are sampled according to the posterior, meaning that, in the limit that the chain length tends to infinity, the relative frequency with which a given set of parameter appears in the chain is proportional to the desired posterior, $p(\vtheta|d,M)$.  Therefore, the output of an MCMC can be directly interpreted as the posterior PDF over the full parameter space, while PDFs for individual parameters can be obtained through trivially marginalizing over the uninteresting parameters, simply by considering the distribution of the parameter of interest in the MCMC samples.

A Markov chain has the property that the probability distribution of the next state can depend only on the current state, not on the past history:
\be
p(\vtheta^{(i+1)})=\int_{V} d\vtheta^{i} p(\vtheta^{(i)} \to \vtheta^{(i+1)}) p(\vtheta^{(i+1)}),
\ee
where $p(\vtheta^{(i)} \to \vtheta^{(i+1)})$ depends only on $\vtheta^{(i)}$ and $\vtheta^{(i+1)}$. 
An additional requirement for an MCMC arises from the fact that the desired distribution is the equilibrium distribution.  In other words, if we assume that state $(i)$ of the chain is sampled from the desired PDF, $p(\vtheta^{(i)})=p(\vtheta|d,M)$ , then the next state $(i+1)$ must be sampled from the PDF as well, so that $p(\vtheta^{(i+1)})=p(\vtheta|d,M)$.

One way to produce such a sequence of samples is via the Metropolis-Hastings algorithm, first proposed by Metropolis et al.~in 1953 \cite{Metropolis:1953}, and later generalized by Hastings:
\begin{enumerate}
  \item Given a current state $\vtheta^{(i)}$, propose the next state $\vtheta^p$ by drawing from a jump proposal distribution with probability $Q(\vtheta^{(i)} \to \vtheta^p)$.  
  \item Compute the probability of accepting the proposed jump as
\bel{eq:p-accept}
p_{\textnormal{accept}}  \equiv \min\Bigl(1,  \frac{p(\vtheta^p|d)}{p(\vtheta^{(i)}|d)} \frac{Q(\vtheta^p \to
        \vtheta^{(i)})}{Q(\vtheta^{(i)} \to \vtheta^p)} \Bigr).
\ee
\item Pick a uniform random number $\alpha \in [0,1]$.  If $\alpha<  p_{\textnormal{accept}}$, accept the proposed jump, setting $\vtheta^{(i+1)}=\vtheta^p$.  Otherwise, reject the jump, and remain at the same location in parameter space for the next step, $\vtheta^{(i+1)}=\vtheta^{(i)}$.
 \end{enumerate}

This jump proposal distribution $Q(\vtheta^{(i)} \to \vtheta^p)$ can depend on the parameters of the current state $\vtheta^{(i)}$, but not on the past history.  It must also allow any state within the allowed prior volume to be reachable by the MCMC.  Beyond these conditions, there is obviously a great deal of freedom in choosing the jump proposal distribution.  This is the most important choice in the MCMC, as it determines the sampling efficiency of the algorithm, i.e., the necessary length of the chain before it converges to the posterior PDF.  Creating an efficient jump proposal distribution requires an understanding of the structure of the parameter space which may not be available until the PDFs are found, creating a Catch-22; one possibility for resolving this infinite loop is described in Section \ref{sec:examples}.

@@

The samples produced during by an MCMC algorithm can be used to
directly perform a Monte Carlo evidence integral.  This results in a
harmonic mean estimator for the evidence, which suffers from a large
variance and a possible bias\will{Ilya: Can you add citations here for
  the suckiness of HM?}.  Additional techniques for the direct
integration of evidence, also based on a kD tree decomposition of the
parameter space (see Sec.~\ref{sec:kDTree}), are described in
\cite{Weinberg2009}.  These techniques are promising, but in some
cases suffer \cite{Farr2010} from large variance and bias as well.  An
alternative approach to model selection among a set of models is based
on performing an MCMC in a ``super-model'' that encompasses all of the
models under consideration; this is known the the Reversible Jump
Markov chain Monte Carlo (RJMCMC).

>>>>

\def\aj{Astronomical Journal}                 % Astronomical Journal
\def\apj{Astrophysical Journal}                % Astrophysical Journal
\def\apjl{Astrophysical Journal}             % Astrophysical Journal, Letters
\def\pasj{PASJ}
\def\apjs{ApJS}              % Astrophysical Journal, Supplement
\def\mnras{MNRAS}            % Monthly Notices of the RAS
\def\prd{Phys.~Rev.~D}       % Physical Review D
\def\prl{Phys.~Rev.~Lett}    % Physical Review Letters
\def\cqg{Class.~Quant.~Grav.~}%Classical and Quantum Gravity
\def\araa{ARA\&A}             % Annual Review of Astron and Astrophys
\def\nat{Nature}              % Nature
\def\aap{A\&A}                % Astronomy and Astrophysics

@@

\begin{abstract}
  Selection among alternative theoretical models given an observed data set is an important challenge in many areas of physics and astronomy, ranging from gravitational-wave data analysis to measurements of black-hole masses in X-ray binaries.  Reversible-jump Markov chain
  monte carlo (RJMCMC) is an extremely powerful technique for
  performing Bayesian model selection, but it suffers from a
  fundamental difficulty: it requires jumps between model parameter
  spaces, but cannot retain a memory of the favored locations in more
  than one parameter space at a time.  Thus, a naive jump between
  parameter spaces is unlikely to be accepted in the MCMC algorithm
  and convergence is correspondingly slow.  Here we demonstrate an
  interpolation technique that uses samples from single-model MCMCs to
  propose inter-model jumps from an approximation to the single-model
  posterior of the target parameter space.  The interpolation
  technique, based on a kD-tree data structure, is adaptive and
  efficient in arbitrary dimensions.  We show that our technique leads
  to dramatically improved convergence over naive jumps in an RJMCMC,
  and compare it to other proposals in the literature to improve the
  convergence of RJMCMCs.  We also discuss the use of the same
  interpolation technique in two other contexts: as a convergence test
  for a single-model MCMC and as a way to construct efficient
  ``global'' proposal distributions for single-model MCMCs without
  prior knowledge of the structure of the posterior distribution.
\end{abstract}

@@

When selecting among alternative models, we are interested in finding
the model with the highest posterior probability $p(M_i|d)$.  However,
attempts to directly compute the evidence by performing the
integration in Eq.~\eqref{evidence} are generally very difficult in a
multi-dimensional, multi-modal parameter space when the likelihood has
to be evaluated numerically.  In particular, a grid-based integral
quickly becomes computationally unfeasible as the dimensionality of
$\vtheta$ exceeds a few.  The parameter space must typically be
explored in a stochastic manner before the evidence integral can be
computed.  There are several stochastic parameter-exploration
techniques focused directly on evidence computation (e.g., nested
sampling \cite{Skilling:2006} and MultiNest \cite{Feroz:2009}).  However, we
frequently want to compute the posterior PDFs within each model along
with the evidences for the various models.  One of the most common
techniques for computing posterior PDFs in the context of a model is
the Markov Chain Monte Carlo, which we now describe.

@@

A Markov chain Monte Carlo \cite{Gilks:1996} produces a set of samples $\{ \vtheta^{(j)} \, | \, i = 1, \ldots \}$ from the model parameter space that are sampled according to the posterior, meaning that, in the limit that the chain length tends to infinity, the relative frequency with which a given set of parameter appears in the chain is proportional to the desired posterior, $p(\vtheta|d,M)$.  Therefore, the output of an MCMC can be directly interpreted as the posterior PDF over the full parameter space, while PDFs for individual parameters can be obtained through trivially marginalizing over the uninteresting parameters, simply by considering the distribution of the parameter of interest in the MCMC samples.

A Markov chain has the property that the probability distribution of the next state can depend only on the current state, not on the past history:
\be
p(\vtheta^{(j+1)})=\int_{V} d\vtheta^{(j)} p(\vtheta^{(j)} \to \vtheta^{(j+1)}) p(\vtheta^{(j)}),
\ee
where the jump probability $p(\vtheta^{(j)} \to \vtheta^{(j+1)})$ depends only on $\vtheta^{(j)}$ and $\vtheta^{(j+1)}$. 
An additional requirement for an MCMC arises from the fact that the desired distribution is the equilibrium distribution.  In other words, if we assume that state $(j)$ of the chain is sampled from the desired PDF, $p(\vtheta^{(j)})=p(\vtheta|d,M)$ , then the next state $(j+1)$ must be sampled from the PDF as well, so that $p(\vtheta^{(j+1)})=p(\vtheta|d,M)$.

One way to produce such a sequence of samples is via the Metropolis-Hastings algorithm, first proposed by Metropolis et al.~in 1953 \cite{Metropolis:1953}, and later generalized by Hastings:
\begin{enumerate}
  \item Given a current state $\vtheta^{(j)}$, propose the next state $\vtheta^p$ by drawing from a jump proposal distribution with probability $Q(\vtheta^{(j)} \to \vtheta^p)$.  
  \item Compute the probability of accepting the proposed jump as
\bel{eq:p-accept}
p_{\textnormal{accept}}  \equiv \min\Bigl(1,  \frac{p(\vtheta^p|d)}{p(\vtheta^{(j)}|d)} \frac{Q(\vtheta^p \to
        \vtheta^{(j)})}{Q(\vtheta^{(j)} \to \vtheta^p)} \Bigr).
\ee
\item Pick a uniform random number $\alpha \in [0,1]$.  If $\alpha<  p_{\textnormal{accept}}$, accept the proposed jump, setting $\vtheta^{(j+1)}=\vtheta^p$.  Otherwise, reject the jump, and remain at the same location in parameter space for the next step, $\vtheta^{(j+1)}=\vtheta^{(j)}$.
 \end{enumerate}

This jump proposal distribution $Q(\vtheta^{(j)} \to \vtheta^p)$ can depend on the parameters of the current state $\vtheta^{(j)}$, but not on the past history.  It must also allow any state within the allowed prior volume to be reachable by the MCMC.  Beyond these conditions, there is obviously a great deal of freedom in choosing the jump proposal distribution.  This is the most important choice in the MCMC, as it determines the sampling efficiency of the algorithm, i.e., the necessary length of the chain before it converges to the posterior PDF.  Creating an efficient jump proposal distribution requires an understanding of the structure of the parameter space which may not be available until the PDFs are found, creating a Catch-22; one possibility for resolving this infinite loop is described in Section \ref{sec:examples}.

@@

The samples produced during by an MCMC algorithm can be used to
directly perform a Monte Carlo evidence integral.  This results in a
harmonic mean estimator for the evidence, which suffers from a large
variance and a possible bias \cite{NewtonRaftery:1994,Chib:1995,vanHaasteren:2009}.  Additional techniques for the direct integration of evidence, also based on a kD tree decomposition of the
parameter space (see Sec.~\ref{sec:kDTree}), are described in
\cite{Weinberg2009}.  These techniques are promising, but in some
cases suffer \cite{Farr2010} from large variance and bias as well.  An
alternative approach to model selection among a set of models is based
on performing an MCMC in a ``super-model'' that encompasses all of the
models under consideration; this is known the the Reversible Jump
Markov chain Monte Carlo (RJMCMC).
